---
title: "Nearly two-thirds of university faculty serve in lower-paying, non-tenured roles"
author: Sapna Bansil
date: "Jan. 13, 2023"
output:
  html_document:
    ##theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Story pitch

As students pay increasingly more for higher education, their instructors are disproportionately likely to work in lower-paying jobs that may be part-time, temporary and/or without benefits.

Nearly two-thirds of faculty members at the University of Maryland held non-tenured positions, an analysis of 2021 salary data revealed. Non-tenured faculty earned an average annual salary of approximately \$71,500, half as much as faculty in tenured or tenure-track positions.

I am interested in writing a reported data story that explores why the university has hired so many non-tenured faculty and contextualizes this finding relative to national trends.

[Source used for this pitch](https://uhr.umd.edu/uhr-support-center/phr-support/general/employment-categories/)

## Libraries required for this analysis

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: true

library(tidyverse)
library(kableExtra)

output_formatted_table <- function(table, text, label_text = NULL){
  table %>%
    kable(caption = text, label = label_text) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 14, fixed_thead = T) %>%
    scroll_box(width = "80%")
}
```

## Loading and cleaning data

I loaded a CSV file with the 2021 salary data and added a new column characterizing each position as tenured or non-tenured based on descriptions from the [University of Maryland Department of Human Resources](https://uhr.umd.edu/uhr-support-center/phr-support/general/employment-categories/).

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: true

salary_data <- read_csv("Data/salary_data.xlsx - Sheet 1.csv")

salary_data <- salary_data %>%
  mutate(position_type = case_when(
    category_status == "Tenured Faculty" | category_status == "Tenure Track Faculty" ~ "tenured",
    TRUE ~ "non-tenured"
  ))

```

## Analysis related to story pitch

To complete this analysis, I first determined the average salary for tenured and non-tenured faculty. Then, I determined the total number of faculty hired to tenured and non-tenured positions. Finally, I joined the two tables to present the findings in one dataframe.

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: true

## Average salary by position type
position_salary <- salary_data %>%
  group_by(position_type) %>%
  summarise(avg_salary = mean(appt_base_annual_salary)) %>%
  arrange(desc(avg_salary))

## Total hires within each position
position_total <- salary_data %>%
  group_by(position_type) %>%
  summarise(total=n()) %>%
  mutate(pct = round(total/sum(total)*100,1)) %>%
  arrange(desc(total))

## Join
position_summary <- position_total %>%
  left_join(position_salary, by = "position_type")

## Output for markdown file
output_formatted_table(position_summary, "Non-tenured versus tenured faculty at Maryland, 2021")


```

