---
title: ""X"
author: Sapna Bansil
date: "Jan. 13, 2023"
output:
  html_document:
    ##theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Story pitch

When the Baltimore Orioles hired Liz Pardo and Kathryn Rowe to their coaching staff, they typified a trend: In an era of increasing gender diversity among coaches in MLB, two-thirds of clubs now employ women in coaching positions at the major or minor league levels. But with the majority of female coaches working in positions such as "mental skills" or "strength and conditioning," it raises a question: How much progress has MLB actually made towards hiring its first female manager?

I am interested in writing a reported data story on trends related to female coaches in MLB, with a spotlight on the Orioles. Using data from [The Institute for Diversity and Ethics in Sports (TIDES)](https://www.tidesport.org/racial-gender-report-card), I created and analyzed a database of female coaches from 2020 to 2022, during which time gender diversity within coaching positions dramatically increased. Among the key findings of my analysis:

* 33 women representing 19 MLB clubs served as coaches in 2022. The Orioles were one of nine teams with at least two female coaches.

* Two-thirds of women who held coaching roles in 2022 worked as mental skills coaches or strength and conditioning coaches, including Pardo and Rowe of the Orioles. Eleven women -- twice as many as in 2020 -- had on-field coaching or player development roles that have traditionally served as a pathway towards an eventual role as an MLB manager.

* The area's other local team, the Washington Nationals, is one of 10 clubs that has not hired a women to any coaching positions over the last three seasons. Since 2020, no women have served for the Orioles in on-field coaching or player development roles.

This analysis is timely given that women, including Miami Marlins general manager Kim Ng, minor league manager Rachel Balkovec and San Francisco Giants coach Alyssa Nakken, are serving in increasingly prominent roles in MLB organizations. It is also unique among other news stories that have analyzed TIDES' data on racial and gender hiring practices in sports because of its breakdown by team and coach type.

```{r echo=FALSE, message=FALSE}

## Libraries required for this analysis
library(tidyverse)
library(janitor)
library(kableExtra)
library(ggplot2)
library(ggthemes)

output_formatted_table <- function(table, text, label_text = NULL){
  table %>%
    kable(caption = text, label = label_text) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 14, fixed_thead = T) %>%
    scroll_box(width = "80%")
}
```

## Source data

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: true

salary_data <- read_csv("Data/salary_data.xlsx - Sheet 1.csv")

```

## Analysis supporting data in memo

**Finding 1:** 33 women representing 19 MLB clubs served as coaches in 2022. The Orioles were one of nine teams with at least two female coaches.

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: true

## URL for description of categories: https://uhr.umd.edu/uhr-support-center/phr-support/general/employment-categories/

## Avg salary by position type
salary_data %>%
  group_by(category_status) %>%
  summarise(avg_salary = mean(appt_base_annual_salary)) %>%
  arrange(desc(avg_salary))

## Total within each position
salary_data %>%
  group_by(category_status) %>%
  summarise(total=n()) %>%
  mutate(pct = round(total/sum(total)*100,1)) %>%
  arrange(desc(total))

## Salaries by school
salary_data %>%
  group_by(division) %>%
  summarise(avg_salary = mean(appt_base_annual_salary)) %>%
  arrange(desc(avg_salary))

## Positions by school
salary_data %>%
  group_by(division, category_status) %>%
  summarise(total = n())

## Schools with most tenure positions
salary_data %>%
  filter(category_status=="Tenured Faculty" | category_status == "Tenure Track Faculty") %>%
  group_by(division) %>%
  summarise(total=n()) %>%
  arrange(desc(total)) %>%
  mutate(pct=round(total/sum(total)*100,1))

```

**Finding 2:** Two-thirds of women who held coaching roles in 2022 worked as mental skills coaches or strength and conditioning coaches, including Pardo and Rowe of the Orioles. Eleven women -- twice as many as in 2020 -- had on-field coaching or player development roles that have traditionally served as a pathway towards an eventual role as an MLB manager.

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: true

## What roles did female coaches in MLB take on between 2020 and 2022?

roles_2022 <- mlb_coaches %>%
  filter(year == "2022") %>%
  group_by(descriptor) %>%
  summarise(roles_2022 = n())

roles_2021 <- mlb_coaches %>%
  filter(year == "2021") %>%
  group_by(descriptor) %>%
  summarise(roles_2021 = n())

roles_2020 <- mlb_coaches %>%
  filter(year == "2020") %>%
  group_by(descriptor) %>%
  summarise(roles_2020 = n())

roles_2022_2021 <- roles_2022 %>%
  left_join(roles_2021, by = "descriptor")

roles_all <- roles_2022_2021 %>%
  left_join(roles_2020, by = "descriptor") %>%
   mutate(roles_2020 = as.numeric(roles_2020)) %>%
  mutate(roles_2020 = case_when(
    is.na(roles_2020) ~ 0,
    TRUE ~ roles_2020
  )) %>%
  arrange(desc(roles_2022)) %>%
  adorn_totals("row")

## For what roles did the Orioles hire female coaches?
orioles_coaches_2022 <- mlb_coaches %>%
  filter(year == "2022") %>%
  filter(team == "Baltimore Orioles") %>%
  select(year, coach, team, descriptor) %>%
  arrange(descriptor)

output_formatted_table(roles_all, "Female Coaches in MLB by Role, 2020-2022")

output_formatted_table(orioles_coaches_2022, "Female Coaches On Baltimore Orioles, 2022")

```

**Finding 3:** The area's other local team, the Washington Nationals, is one of 10 clubs that has not hired a women to any coaching positions over the last three seasons. Since 2020, no women have served for the Orioles in on-field coaching or player development roles.

```{r}
#| echo: true
#| warning: false
#| message: false
#| code-fold: true

## Which teams have not hired any female coaches?
mlb_2020_to_2022 <- mlb_coaches %>%
  group_by(team) %>%
  summarise(female_coaches = n())

all_mlb <- mlb_teams %>%
  left_join(mlb_2020_to_2022, by = "team") %>%
   mutate(female_coaches = as.numeric(female_coaches)) %>%
  mutate(female_coaches = case_when(
    is.na(female_coaches) ~ 0,
    TRUE ~ female_coaches
  )) %>%
  mutate(had_female_coaches = case_when(
    female_coaches == 0 ~ "no",
    female_coaches >= 1 ~ "yes"
  ))

teams_without_coach <- all_mlb %>%
  filter(female_coaches == "0") %>%
  select(team, female_coaches)

## Which teams have not hired female coaches to on-field or player development roles
mlb_on_field <- mlb_coaches %>%
  filter(descriptor == "On-Field Coaching" | descriptor == "Player Development") %>%
  group_by(team) %>%
  summarise(on_field_coaches = n()) %>%
  arrange(team)

all_mlb_on_field <- mlb_teams %>%
  left_join(mlb_on_field, by = "team") %>%
   mutate(on_field_coaches = as.numeric(on_field_coaches)) %>%
  mutate(on_field_coaches = case_when(
    is.na(on_field_coaches) ~ 0,
    TRUE ~ on_field_coaches
  )) %>%
  mutate(had_on_field_coaches = case_when(
    on_field_coaches == 0 ~ "no",
    on_field_coaches >= 1 ~ "yes"
  ))

teams_without_on_field_coach <- all_mlb_on_field %>%
  filter(on_field_coaches == "0") %>%
  select(team, on_field_coaches)

output_formatted_table(teams_without_coach, "MLB Teams Without Female Coaches, 2020-2022")

output_formatted_table(teams_without_on_field_coach, "MLB Teams Without Female Coaches in On-Field or Player Development Roles, 2020-2022")

```
